---
title: "House Rent Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | House Rent Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/iamsouravbanerjee/house-rent-prediction-dataset?select=House_Rent_Dataset.csv\>*

### Reference:

*\<Banerjee, S. (n.d.). House Rent Prediction Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/iamsouravbanerjee/house-rent-prediction-dataset?select=House_Rent_Dataset.csv\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r loading dataset}
# Load the dataset
HouseRentData <- read.csv("House_Rent_Dataset.csv", colClasses = c(
  BHK = "numeric",                     # Number of bedrooms
  Rent = "numeric",                    # Rent amount
  Size = "numeric",                    # Size in square feet
  Area_Type = "factor",                # Type of area (e.g., Super Area)
  Area_Locality = "factor",         # Locality name
  City = "factor",                     # City name
  Furnishing_Status = "factor",        # Furnishing status (e.g., Furnished)
  Tenant_Preferred = "factor",         # Tenant preference
  Bathroom = "numeric",                # Number of bathrooms
  Point_of_Contact = "factor"       # Point of contact information
))

# Display structure to verify data types
str(HouseRentData)

# Display first few rows to ensure data is loaded correctly
head(HouseRentData)

# Optional: View the dataset in spreadsheet-like interface
View(HouseRentData)
```

## Measures of Frequency
```{r MOF}
# Load necessary libraries
library(dplyr)  # For data manipulation
library(ggplot2)  # For visualizations
library(corrplot)  # For correlation plots

# Assuming `HouseRentData` is already loaded

# Measures of Frequency
# Frequency distribution for categorical variables
frequency_analysis <- function(data, column_name) {
  return(data %>%
           group_by(!!sym(column_name)) %>%
           summarize(Frequency = n()) %>%
           arrange(desc(Frequency)))
}

# Example: Frequency of cities
city_frequency <- frequency_analysis(HouseRentData, "City")
print(city_frequency)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean, Median, Mode
mean_rent <- mean(HouseRentData$Rent, na.rm = TRUE)
median_rent <- median(HouseRentData$Rent, na.rm = TRUE)
mode_rent <- as.numeric(names(sort(table(HouseRentData$Rent), decreasing = TRUE)[1]))

cat("Mean Rent:", mean_rent, "\n")
cat("Median Rent:", median_rent, "\n")
cat("Mode Rent:", mode_rent, "\n")
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Range, Variance, Standard Deviation, Interquartile Range
range_rent <- range(HouseRentData$Rent, na.rm = TRUE)
variance_rent <- var(HouseRentData$Rent, na.rm = TRUE)
sd_rent <- sd(HouseRentData$Rent, na.rm = TRUE)
iqr_rent <- IQR(HouseRentData$Rent, na.rm = TRUE)

cat("Range of Rent:", range_rent, "\n")
cat("Variance of Rent:", variance_rent, "\n")
cat("Standard Deviation of Rent:", sd_rent, "\n")
cat("Interquartile Range of Rent:", iqr_rent, "\n")
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation Matrix
# Selecting numeric variables for correlation
numeric_columns <- HouseRentData %>%
  select_if(is.numeric)

correlation_matrix <- cor(numeric_columns, use = "complete.obs")
print(correlation_matrix)

# Correlation heatmap
corrplot(correlation_matrix, method = "circle", type = "upper", tl.col = "black")

# Example: Scatter plot to visualize relationship
ggplot(HouseRentData, aes(x = Size, y = Rent)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Relationship between Size and Rent",
       x = "Size (sq ft)",
       y = "Rent (₹)")
```

## ANOVA
```{r ANOVA}
# Check structure of the dataset
str(HouseRentData)

# Perform one-way ANOVA
anova_result <- aov(Rent ~ Furnishing_Status, data = HouseRentData)

# Display ANOVA summary
summary(anova_result)

# Post-hoc test (if ANOVA is significant, use Tukey's HSD to determine pairwise differences)
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)
```

## Plots
```{r plots}
# Visualize group means
ggplot(HouseRentData, aes(x = Furnishing_Status, y = Rent, fill = Furnishing_Status)) +
  geom_boxplot() +
  labs(title = "Boxplot of Rent by Furnishing Status",
       x = "Furnishing Status",
       y = "Rent") +
  theme_minimal()

# Load necessary libraries
library(ggplot2)
library(corrplot)
library(dplyr)

# Univariate Plots

# Histogram for numeric variable (e.g., Rent)
ggplot(HouseRentData, aes(x = Rent)) +
  geom_histogram(binwidth = 5000, fill = "blue", color = "black") +
  labs(title = "Histogram of Rent", x = "Rent (₹)", y = "Frequency") +
  theme_minimal()

# Boxplot for numeric variable (e.g., Size)
ggplot(HouseRentData, aes(y = Size)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Boxplot of House Size", y = "Size (sq ft)") +
  theme_minimal()

# Bar plot for categorical variable (e.g., Furnishing_Status)
ggplot(HouseRentData, aes(x = Furnishing_Status)) +
  geom_bar(fill = "purple", color = "black") +
  labs(title = "Bar Plot of Furnishing Status", x = "Furnishing Status", y = "Count") +
  theme_minimal()

# Multivariate Plots

# Scatter plot for two numeric variables (e.g., Size vs. Rent)
ggplot(HouseRentData, aes(x = Size, y = Rent)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Scatter Plot of Size vs Rent", x = "Size (sq ft)", y = "Rent (₹)") +
  theme_minimal()

# Boxplot for numeric variable grouped by categorical variable (e.g., Rent by Furnishing_Status)
ggplot(HouseRentData, aes(x = Furnishing_Status, y = Rent, fill = Furnishing_Status)) +
  geom_boxplot() +
  labs(title = "Boxplot of Rent by Furnishing Status", x = "Furnishing Status", y = "Rent (₹)") +
  theme_minimal()

# Faceted plots to compare distributions (e.g., Rent distribution by City)
ggplot(HouseRentData, aes(x = Rent)) +
  geom_histogram(binwidth = 5000, fill = "green", color = "black") +
  facet_wrap(~City) +
  labs(title = "Histogram of Rent by City", x = "Rent (₹)", y = "Frequency") +
  theme_minimal()

# Heatmap of correlation matrix (numeric variables)
numeric_columns <- HouseRentData %>%
  select_if(is.numeric)
correlation_matrix <- cor(numeric_columns, use = "complete.obs")
corrplot(correlation_matrix, method = "color", type = "upper", tl.col = "black")


```

# Preprocessing & Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in the dataset

# Summarize missing values for each column
missing_summary <- colSums(is.na(HouseRentData))

# Display the summary of missing values
print(missing_summary)

# Check if there are any missing values in the entire dataset
any_missing <- any(is.na(HouseRentData))
cat("Are there any missing values in the dataset? ", any_missing, "\n")

# Visualize missing values (optional, requires the 'naniar' library)
library(naniar)

# Plot missing values
gg_miss_var(HouseRentData) +
  labs(title = "Missing Values by Variable",
       x = "Variables", y = "Number of Missing Values") +
  theme_minimal()

```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load necessary library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Data Splitting: Split data into training (70%) and testing (30%)
trainIndex <- createDataPartition(HouseRentData$Rent, p = 0.7, list = FALSE)
trainData <- HouseRentData[trainIndex, ]
testData <- HouseRentData[-trainIndex, ]
cat("Training data rows:", nrow(trainData), "\n")
cat("Testing data rows:", nrow(testData), "\n")
```

## Bootstrapping
```{r Bootstrapping}
# Bootstrapping: Create bootstrapped samples
bootstrapSamples <- createResample(HouseRentData$Rent, times = 1000, list = TRUE)
bootstrapMeans <- sapply(bootstrapSamples, function(indices) {
  mean(HouseRentData$Rent[indices], na.rm = TRUE)
})
cat("Bootstrap Means Summary:\n")
print(summary(bootstrapMeans))
```

## Cross-validation
```{r Cross-validation}
# Basic Cross-Validation: 5-fold CV
trainControlCV <- trainControl(method = "cv", number = 5)
cvModel <- train(Rent ~ ., data = trainData, method = "lm", trControl = trainControlCV)
cat("Basic CV Model Summary:\n")
print(cvModel)

```

## Training Different Models
```{r Different Models}
# Load necessary libraries
library(caret)
library(glmnet)
library(randomForest)

# Set seed for reproducibility
set.seed(123)

# Linear Regression (LM)
lm_model <- train(Rent ~ ., data = trainData, method = "lm")
cat("Linear Regression Model Summary:\n")
print(lm_model)

# Ridge Regression (using glmnet)
ridge_model <- train(Rent ~ ., data = trainData, method = "glmnet", 
                     tuneGrid = expand.grid(alpha = 0, lambda = seq(0.001, 0.1, by = 0.001)))
cat("Ridge Regression Model Summary:\n")
print(ridge_model)

# Random Forest Regression (using randomForest package)
rf_model <- randomForest(Rent ~ ., data = trainData)
cat("Random Forest Regression Model Summary:\n")
print(rf_model)

# Optional: Predicting using the models on the test set
lm_predictions <- predict(lm_model, newdata = testData)
ridge_predictions <- predict(ridge_model, newdata = testData)
rf_predictions <- predict(rf_model, newdata = testData)

# Evaluate performance of models (using RMSE as example)
lm_rmse <- sqrt(mean((lm_predictions - testData$Rent)^2))
ridge_rmse <- sqrt(mean((ridge_predictions - testData$Rent)^2))
rf_rmse <- sqrt(mean((rf_predictions - testData$Rent)^2))

cat("Linear Regression RMSE:", lm_rmse, "\n")
cat("Ridge Regression RMSE:", ridge_rmse, "\n")
cat("Random Forest Regression RMSE:", rf_rmse, "\n")
```

## Performance Comparison with resamples
```{r Performance Comparison}
# Combine the models into a list for resampling
model_list <- list(LM = lm_model, Ridge = ridge_model, RF = rf_model)

# Resampling to compare the models
resamples_results <- resamples(model_list)

# Summary of resamples for performance comparison
summary(resamples_results)

# Boxplot to visually compare the model performances
bwplot(resamples_results, metric = "RMSE") # Boxplot of RMSE for all models
bwplot(resamples_results, metric = "Rsquared") # Boxplot of R-squared for all models
```

## Saving Model
```{r Saving Model}
# Saving the Ridge Regression model
saveRDS(ridge_model, "./models/saved_ridge_model.rds")

# Load the saved Ridge Regression model
loaded_ridge_model <- readRDS("./models/saved_ridge_model.rds")

# Example new data for prediction (adjust based on your features)
new_data <- data.frame(
  BHK = 2,                       # Number of bedrooms
  Size = 1000,                    # Size in square feet
  Floor = "1 out of 3",           # Floor information
  Area_Type = "Super Area",       # Type of area
  Area_Locality = "Salt Lake City", # Locality name
  City = "Kolkata",               # City name
  Furnishing_Status = "Semi-Furnished", # Furnishing status
  Tenant_Preferred = "Bachelors/Family", # Tenant preference
  Bathroom = 1,                    # Number of bathrooms
  Point_of_Contact = "Contact Owner" # Point of contact (ensure this is included)
  
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_ridge_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

